---
title: "Chi-Square Examples"
output: html_document
---
  
```{r knitr_init, echo=FALSE, cache=FALSE, warning=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```
## Load `tidyverse` package

We will use tidyverse functions for all these examples.

```{r}
# load packages -----------------------------------------------------------
library(tidyverse)  # load tidyverse, for working with datasets
```

## Proportional Model Example

### Goal

Replicate the analysis done in Example 8.1 in The Analysis of Biologica Data (p. 204).

Answer the question: **Do births occur randomly throughout the week?**

H0: the number of births does *not* vary by day of the week.
Ha: the number of births *does* vary by day of the week.

### Data

Use the data from the Births dataset from the mosaicData package:
https://vincentarelbundock.github.io/Rdatasets/doc/mosaicData/Births.html

```{r}
# load the mosaicData package
library(mosaicData)

# put the dataset in the global environment as a tibble and rename it to lowercase
births <- as_tibble(Births)
births
```

The dataset contains the number of births each day from 1968 to 1988. 

It is a data frame with the following 8 variables:

- `date` Date
- `births` Number of births on date (integer)
- `wday` Day of week (ordered factor)
- `year` Year (integer)
- `month` Month (integer)
- `day_of_year` Day of year (integer)
- `day_of_month` Day of month (integer)
- `day_of_week` Day of week (integer)

Take a look at the data:

```{r, echo=FALSE}
DT::datatable(births)
```


### Analysis

#### Summarize data by day of week

The goal is to get a table with day of week as a variable (`yday`) and a
variable for "number of births".

The `births` variable already has the number of births on a given date. We just
have to sum those up for each day of the week.

```{r}
# create frequency table
ft <- births %>% 
  group_by(wday) %>% 
  summarize(births = sum(births))
```

Now to visualize the data.

If we were using the raw data to make this plot, we would use a barchart, which would do the counting for us.

Because we have already counted the births, we will use a column chart. 

```{r}
# plot 
ggplot(ft) +
  geom_col(aes(x = wday, y = births)) +
  labs(
    x = "Day of the week",
    y = "Frequency",
    title = "Births 1968-1988"
  )
```

### Conduct GOF test

```{r}
# calcualte expected frequencies
expected <-
  ft %>% 
  mutate(
    total = sum(births),
    n_cat = n(),
    expected = total / n_cat
  )
expected
```

#### Calculate chi-square test statistic and interpret

```{r}
# calcualte the chi-square for each row
chi_sq_table <-
  expected %>% 
  mutate(
    chi_sq = ( (births - expected)^2 ) / expected
  )
chi_sq_table
```

Add up the chi-square from each row to get the chi-square test statistic.

At the same time, count the categories and use that to calculate the degrees of freedom (df) and the critical value.

```{r}
chi_sq_table %>% 
  summarize(
    chi_sq_test_stat = sum(chi_sq),
    n_cat = n(),
    df = n_cat - 1,
    alpha = 0.05,
    crit_val = qchisq(p = 1 - alpha, 
                      df = df)
  )
```

Is the chi-square test statistic greater than the critical value?

If yes, then reject the null hypothesis and accept the alternative hypothesis.

Ha: The number of births *does* vary by day of the week.

### Alternative way to do it

You can also use the `chisq.test` function in R.

To do this, you will a vector of the observered frequencies.

```{r}
# observered frequencies
(x <- ft$births)
```

Use the `chisq.test` function

```{r}
(Xsq <- chisq.test(x))
Xsq$observed   # observed counts (same as x)
Xsq$expected   # expected counts under the null
Xsq$statistic  # value the chi-squared test statistic
Xsq$p.value    # p-value for the test
```
Instead of comparing the test statistic to a critical value, you compare the
p-value to your alpha level.

In this case the p-value of 2.2e-16 (0.00000000000000022) is less than our alpha
value of 0.05 so we reject the null hypothesis.

Note: this assumes the expected proportion for each category is `1/n_cat`. Or 
1/7 in our case.

If the expected proportion is different, then you would need to provide a second 
argument `p` that is the same length as `x` which contains the expected 
proportions.

If you have something other than expected proportions that does not sum to 1,
for example if you have expected frequencies, then you can use those for `p`
and add the argument `rescale.p = TRUE`, which will rescale `p` so that it sums
to 1. It does this by dividing each element of `p` by the sum of `p`.

## Poisson Example

## Contingency Test Example

This is an example of how to conduct the chi-square contingency test shown in
Example 9.4 in your textbook (p. 246).

While it is possible to use `chisq.test()` to do this, that function requires
the data to be in a **matrix** data type. In this class, we mostly use 
**data frames**, and in particular **tibbles**.

Here's how to convert from a data frame to a tibble:

```{r}
library(infer)
```

Let's start with our contingency table:

```{r}
fish <-
  tribble(
            ~infection,      ~eaten, ~n_fish,
          "Uninfected",     "Eaten",       1,
    "Lightly infected",     "Eaten",      10,
     "Highly infected",     "Eaten",      37,
          "Uninfected", "Not Eaten",      49,
    "Lightly infected", "Not Eaten",      35,
     "Highly infected", "Not Eaten",       9
  )
```

Now using the infer package you can do the chi-square goodness-of-fit test like this:

```{r}

(fish_matrix <- matrix(fish$n_fish, nrow = 2, byrow = TRUE))
chisq.test(fish_matrix)

```

#### Using raw data

What if your data frame contained the raw data, i.e. 141 rows with each row
representing a single fish.

Luckily, where there is a need, there is usually a package. Users have created
the **infer** package to facilitate performing statistical inference using the
same grammar as the **tidyverse**.

```{r, eval=FALSE}
# generate the raw data
fish_observations <-
  fish %>% 
  uncount(n_fish) %>% 
  print()

# do the chi-square test
fish_observations %>%
  chisq_test(response = eaten, explanatory = infection)
```

