# (PART) EDA Project {.unnumbered}

# EDA Project

## Overview

A major part of the lab will be to perform your own exploratory data analysis in R.

You will choose a dataset to use and a question to answer. You will present your results to the class and at the [MSUM's Andrew B. Conteh Student Academic Conference (SAC)](https://www.mnstate.edu/sac/) in April (application deadline is Feb. 25).

## Purpose

The purpose of the lab project is for you to practice working with real data, exploring data sets to answer a question that interests *you*, and communicating the results.

## Choose a project

### Data sets

The first step in your EDA project is to choose one or more data sets you will explore.

For help in finding, learning about, and accessing data sets, see [Data Sets](data-sets.html).

For some questions, you may need to use multiple datasets. For example you may combine county level health data with county level environmental data from different datasets.

### Question

The second step in your EDA project is to develop a broad question or concept that can be answered or explored with your data sets.

For example, if you are interested in the field of public health, your over-arching question might be something like "How do environmental risk factors affect cardiovascular diseases?"

During the course of your exploratory data analysis you will gradually form more specific questions. These questions are will be phrased in terms of the variables and observations that comprise your data sets, and each will have a definite answer in the form of a visualization or statistical analysis. Following the diabetes example, your more specific questions might include the following:

-   How did diabetes prevalence vary among states in 2019?
-   How did diabetes prevalence vary among racial or ethnic groups in Minnesota in 2019?
-   What is the correlation between annual diabetes prevalence and the proportion of the population covered by medical insurance in Clay County, Minnesota from 2010-2019?

Your general question will provide the overall direction of your project, while new specific questions will arise through the course of the exploratory analysis.

You may not end up including all specific questions in your final EDA Project Poster, depending on how interesting or applicable the results were for each one.

## Assignments

1.  [SAC Application](sac-application.html) - Submit your application to the SAC website by Sunday, February 28, 2021

2.  Preliminary Methods and Results

    a.  Add a basic outline for your README, including level 2 headings for **Abstract**, **Introduction**, **Methods**, **Results**, **Discussion**, **Literature Cited**, and **References**.

    b.  Add level 3 headings under Methods for the following sections:

        i.  **Data Acquisition**. In paragraph or list form, describe how you obtained the data. Be specific... which websites did you go to, did you select which data to download, did you have to request access, etc. Use in-text citations to cite your data sources... find on their websites how to cite them appropriately. Put full references in your literature cite.

        ii. Data Preparation. In paragraph or list form, describe how you cleaned up the data in preparation for your analyses.

    c.  Request peer review by **March 25**; complete peer review and respond to changes by **March 30**. Submit to the D2L assignment the link to the peer review of your own project by March 30. You will be graded on your methods section and on the quality of your the peer review(s) you provide to others.

## References

You should have a section titled "References" containing a list of references you cite in your poster.

Entries in the References section are commonly called "end citations", while citations in the text of your poster are called "in-text citations". Each each in-text citation should have a corresponding end citation and vice versa.

There are three types of references you may cite:

1.  **Primary literature**. Published peer-reviewed journal articles. Articles from the popular press (e.g. a magazine, blog, or news website) should not be used.
2.  **Data sources**. Where you obtained your data from.
3.  **Software**. Including R, RStudio, and any R packages you used.

### Citing software

R makes it easy to cite software by giving you the citations to use. You can find full citations for R using:

```{r}
citation()
```

Always give the version of R and RStudio you are using. In the text of your methods, you would say something like "I used R Version 4.0.3 (R Core Team 2021)"

To find your version of R type:

```{r}
R.Version()
```

The citation and version number for RStudio can be found like this:

```{r}
RStudio.Version()
```

Citations for R packages can be found like this:

```{r}
citation("dplyr")
```

You should any non-base packages you use. For example, cite any of the tidyverse package (tibble, dplyr, tidyr, forcats, lubridate, readr), but do not cite the tidyverse package itself.

If you use a special package for reading in data, cite is, e.g. readxl.

If you read data from online using a special package, cite it too, e.g. rgbif, rinat, BIEN, auk.

### End citation format

Use the format defined by the Council of Science Editors.

### In-text citation format

In-text citations should be given in the Name-Year format. Examples:

1.  One author: (Merkord 2021)
2.  Two authors: (Merkord and Merkord 2021)
3.  Three authors: (Merkord et al. 2021)

## Project Management

### GitHub repository

Your EDA project will have its own GitHub repository which will include:

1.  Your data analysis R scripts

2.  README.Rmd containing the text and code to create your final poster

3.  README.md containing your final poster. This document is what you will present at the SAC

### Data Import Tips

This info will be useful when you get to the stage of importing data (after your SAC abstract has been submitted).

Where to store your data?

-   ALWAYS in your project folder, never elsewhere on your computer

-   Create a folder named `data`

-   Add `data/` to your gitignore file

-   download data files and save them into your `data/` folder

    -   you can download manually (point and click in browser). if so, write the link to the data source in your README so its easy to find
    -   you can download programatically with `download.file()` function

Importing your data

-   never edit your raw data file. instead, read it as is and clean it up in R

-   use read_csv() or readxl::read_excel()

    -   do NOT use read.csv()
    -   readxl package is installed with tidyverse, but not loaded
    -   you have to load it yourself: library(readxl)
    -   path argument: type "" put cursor in between and hit tab to autocomplete

-   read_excel() has a `sheet` argument to choose which sheet

-   R puts back ticks around non-standard column names eg `-` is a backtick, its on the upper left of the keyboard

    -   standard names have only letters, numbers, `_` and `.`
    -   any name with special characters or starts with a number is considered non-standard.
    -   its a good idea to rename these columns

-   You can rename columns during import with the col_names argument to `read_*()` or using `dplyr::rename()`

-   You can skip rows using the skip argument to `read_*()`

-   When you play with `skip` argument, sometimes you need to change the `col_names` argument too

-   You can read only a set number of rows using the `n_max` argument
