---
title: "Lab 5: The normal distribution"
output: html_document
---
  
```{r knitr_init, echo=FALSE, cache=FALSE, warning=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

Background reading: Chapter 10 of Whitlock & Schluter (2018).

## Objectives

Learn how to:

1. [Visualize a normal distribution](#visualize-a-normal-distribution)
2. [Convert a variable to *Z*-scores](#convert-a-variable-to-z-scores)

## Instructions

1. Open RStudio
2. Create a new R script for this lab and save it with a name you will recognize later.
3. Copy the [Template](#template) at the end of this lab and paste it into your script.
4. Work through the [Practice](#practice) section, running the code provided.
5. Write your own code and comments to complete the [Assignment](#assignment).
6. Save your R script.
7. Go to the Lab 5 assignment in D2L.
8. Submit your R script. 
    - Make sure the R script is saved in RStudio before submitting.
    - Make sure you select the R script not the Rproj file (the blue cube icon)
    - After submitting, view your submission to make sure it is correct and complete

## Practice

### 1. Visualize a normal distribution

For our first example, let's generate a random sample of 100 measurements from a 
population that has a normal distribution with a mean of 13.5 and a standard 
deviation of 2.82 and assign that sample to a variabled named `x`.

```{r}
# randomly sample a normal distribution
set.seed(52)
x <- rnorm(n = 1000, mean = 7.5, sd = 2.82)
```

You might be wondering about the `set.seed()` function. That ensures that the
random numbers you generate when running this code will be the same as the ones
shown here. YOU MUST RUN `set.seed()` immediately before any random number
generator like `rnorm()` to in order to get the same random numbers. If you 
re-run the code later, always be sure to run `set.seed()` before generating your
vector of values for `x`.

Now print `x` to see the numbers you generated. 

```{r, eval=FALSE}
x    # print x in the console
```

Now let's see how well the sample represents the population it was drawn from by
calculating the mean and standard deviation of the sample.

```{r}
mean(x)    # calculate the mean
sd(x)    # calculate the standard deviation
```
NOTE: If you get any numbers for the mean and standard deviation other than those
shown above, then it is probably because you didn't generate the same set of
random numbers as we did here. Re-run the `set.seed()` and `x<-rnorm()` lines 
above, in that order.

Those numbers are fairly close to the population parameters we used to generate
the data (mean = 7.5, sd = 2.82).

But does the sample have a normal distribution? Let's graph it to make sure.
Remember that an effective way to visualize a single variable is with a 
histogram.

As always, we will use `ggplot()` for graphing, which means we need to load the
**tidverse** package. It also means we need to put x in a data frame, because
the data argument to `ggplot()` must always be a data frame (not a vector). Note
the use of `geom_vline()` to add a vertical line for the mean value.

```{r, message=FALSE}
# load the tidyverse package for ggplot2, tibble, and dplyr
library(tidyverse)

# create a data frame (tibble) to use for plotting
df <- tibble(x)
df
```

```{r plot-histogram}
# plot a histogram of x
ggplot(df, aes(x = x)) +
  geom_histogram(bins = 11) +
  geom_vline(aes(xintercept = mean(x)), color = "darkred")
```

Looking at the histogram, you might think that its not particularly bell-shaped
with that seemingly left-skewed distribution. This kind of outcome is 
actually common when taking a random sample. Those outlier values below -1 are
fairly rare and only represent a tiny proportion of the sample. If you sampled
the same distribution again, you could just as easily have outliers on the other
end of the distribution, or on both, or no outliers at all. The overall
shape and center of gravity, however, would be the same.

### 2. Convert a variable to *Z*-scores

To convert a variable to Z-scores, you subtract its mean and divide by its
standard deviation. This is called ***standardizing*** a variable, or sometimes 
*scaling* (see note below). The resulting numbers are called with 
***Z*-scores**.

Remember, the Z distribution has a mean of 0 and a standard deviation of 1. Our
variable has a mean of about 7.39. 

To get our variable to have a mean of zero, we would need to subtract 7.39 from 
each value. This is called *centering.*

```{r, echo=FALSE}
# center x
x_centered <- x - mean(x)

x_centered[1:20]    # print the first 20 values
```

Let's check if the mean of the new variable is zero.

```{r}
mean(x_centered)    # calculate the mean
```

That's very close to zero. The "e-16" is exponential notation meaning the actual
mean is about 0.0000000000000001067798, or just about zero. Why is it not zero?
Basically it's due to how R rounds very small numbers.

Our variable `x` is centered now, but it still has a standard deviation of 2.87.

```{r}
sd(x_centered)    # calculate the standard deviation
```

The next step is to *scale* the variable so that it has a standard deviation of 
1. This can be done by dividing each value by the variable's standard deviation 
of 2.87.

```{r}
# scale the variable
x_scaled <- x_centered / sd(x_centered)
sd(x_scaled)
mean(x_scaled)
```

That's it... you have now scaled the variable `x`. The values f `x_scaled` are
now called "*Z*-scores". They have a mean of 0 and a standard deviation of 1.

```{r plot-z-scores}
# plot a histogram of x_scaled
ggplot(tibble(x_scaled), aes(x = x_scaled)) +
  geom_histogram(bins = 11) +
  geom_vline(aes(xintercept = mean(x_scaled)), color = "darkred")
```

A note on terminology:

Converting a variable to Z-scores is often called "standardizing", but it is
also sometimes refered to as "scaling". This can be confusing because "scaling"
can thus refer in the loose sense to the act of standardizing, or in the strict
sense to dividing by the standard deviation.

On a related note, there is actually a handy function in R that both centers
and scales (in the strict sense) called. Confusingly the function is called
`scale()`.

Now you know how to standardize a variable stored as a vector. But what if you
want to standardize a variable in a data frame? With `dplyr` it's super easy!

```{r}
# start with data frame and then
# create a new variable for z-scores
mutate(df, z = (x - mean(x)) / sd(x))
```

## Assignment

For the assignmet we will use the `heights` data set in the dslabs package.

The data set contains the height and sex of 1,050 students.

First, install **dslabs** using the Packages tab in RStudio. Then load the `heights`
data set and turn it into a tibble so it prints nicely.

```{r}
library(dslabs)
data("heights")
h <- as_tibble(heights)
h
```

Your assignment is to:

1. Plot the distribution of heights for all students. Use any type of graph that 
seems appropriate.
2. Add a comment to your script stating whether the distribution looks normal or 
not.
3. Standardize the variable `height` (for all students, not females and males
separately).
4. Plot the distribution of Z-scores. Use different aesthetics or facets for 
male and female students to visualize each group seperately but on the same 
graph. Review 
[R4DS Chapter 3 Data Visualisation](https://r4ds.had.co.nz/data-visualisation.html) 
if needed. If you can't figure out how to do it, create two separate plots, one 
for female and one for male, for partial credit.

## Template

Copy and paste this template into your R script.

Fill in the missing elements under the "assignment" heading

```{r, eval=FALSE}
#### Lab 5 Script Template

### Practice ----------------------------------------------

## Objective 1. Visualize a normal distribution -----------

# randomly sample a normal distribution
set.seed(52)
x <- rnorm(n = 1000, mean = 7.5, sd = 2.82)

x             # print x in the console
mean(x)       # calculate the mean
sd(x)         # calculate the standard deviation

# load the tidyverse package for ggplot2, tibble, and dplyr
library(tidyverse)

# create a data frame (tibble) to use for plotting
df <- tibble(x)
df

# plot a histogram of x
ggplot(df, aes(x = x)) +
  geom_histogram(bins = 11) +
  geom_vline(aes(xintercept = mean(x)), color = "darkred")


## Objective 2. Convert a variable to *Z*-scores ----------

# center x
x_centered <- x - mean(x)

x_centered[1:20]    # print the first 20 values
mean(x_centered)    # calculate the mean
sd(x_centered)    # calculate the standard deviation

# scale the variable
x_scaled <- x_centered / sd(x_centered)
sd(x_scaled)
mean(x_scaled)

# plot a histogram of x_scaled
ggplot(tibble(x_scaled), aes(x = x_scaled)) +
  geom_histogram(bins = 11) +
  geom_vline(aes(xintercept = mean(x_scaled)), color = "darkred")

# start with data frame and then
# create a new variable for z-scores
mutate(df, z = (x - mean(x)) / sd(x))


### Assignment -----------------------------------------
### ENTER YOUR CODE BELOW

# 1. Plot the distribution of heights for all students. 



# 2. Does the distribution look normal?  (put 'yes' or 'no' here)



# 3. Standardize the variable `height`.



# 4. Plot the distribution of Z-scores for males and females. Use different
#    aesthetics or facets for sex, or create separate plots for partial credit.



```

