---
title: "Lab 6: Comparing means / Lab Practical"
output: html_document
---
  
```{r knitr_init, echo=FALSE, cache=FALSE, warning=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               cache=TRUE,
               prompt=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

Background reading: Chapters 12 of Whitlock & Schluter (2018).

Lab 6 and the lab practical have been combined so you only have to do one
assignment before the end of the semester.

## Objectives

Learn how to:

1. [Conduct a t-test](#conduct-a-t-test)

Demonstrate your ability to:

1. Wrangle data
2. Plot data
3. Estimate a population parameter
4. Conduct a hypothesis test

## Instructions

1. Open RStudio
2. Create a new R script for this lab and save it with a name you will recognize later.
3. Copy the [Template](#template) at the end of this lab and paste it into your script.
4. Work through the [Practice](#practice) section, running the code provided.
5. Write your own code and comments to complete the [Assignment](#assignment).
6. Save your R script.
7. Go to the *Lab 6 / Lab Practical* assignment in D2L.
8. Submit your R script. 
    - Make sure the R script is saved in RStudio before submitting.
    - Make sure you select the R script not the Rproj file (the blue cube icon)
    - After submitting, view your submission to make sure it is correct and complete

## Practice

### 1. Conduct a t-test

For this example, we will use the iris dataset in the package datasets. This
packaged is loaded automatically when R starts, so you already have access to
it.

Print the iris dataset in the console to see what it looks like:

```{r, eval=FALSE}
# print iris
iris
```

You will notice that it does not have the pretty display of a tibble, but the
clunky display of a base R data frame.

To make it easier to read, convert it to a tibble. In this case, we can just
write over the old iris dataset with the new one.

```{r, message=FALSE}
# turn iris into a tibble
library(tidyverse)
iris <- as_tibble(iris)
iris
```

Notice that "iris" now shows up in your environment tab.

To remind yourself what's in the iris dataset, you can do some basic information
gathering:

```{r}
# look at the dataset
help("iris")
glimpse(iris)
count(iris, Species)
```

As you can see, iris has 150 observatons (rows) and 5 variables (columns). Each 
observation represents an individual plant. The **Species** variable has three 
possible values, one for each species. Each species has 50 observations.

If we wanted to graph the distribution of a variable, for example petal length,
among the three species we could do something like this:

```{r petal-length, fig.height=4, fig.width=6}
# density plot of petal lengths by species
ggplot(iris, aes(x = Petal.Length, color = Species)) +
  geom_density()
```

Sometimes, however, we only want to plot a subset of the data. For example, if
we wanted a histogram of sepal lengths just for species *setosa*, we could
filter the data before plotting:

```{r setosa-hist, fig.width=4, fig.height=3}
# filter setosa data
setosa_data <- filter(iris, Species == "setosa")

# histogram of petal lengths for setosa
ggplot(setosa_data, aes(x = Petal.Length)) +
  geom_histogram(bins = 9)
```

Let's say we wanted to go a step further an calculate the mean petal length.

For this we can use the `mean()` function. The only catch is that the argument
for the `mean()` function must be a vector (a series of numbers), which we do
not have. Instead we have a data frame `setosa_data` with a variable.

To extract the variable as a vector we can use the `pull()` function from 
**dplyr**:

```{r}
# extract petal lengths for setosa
setosa_petal_length <- pull(setosa_data, Petal.Length)
setosa_petal_length

# calculate mean petal length for setosa
mean(setosa_petal_length)
```

This ability becomes very useful for a number of statistics functions in base R
that require vectors not data frames as arguments. For example, the `t.test()`
function.

Let's test whether the mean petal length for *setosa* flowers is different from
the mean petal length for *versicolor* flowers.

`t.test()` requires two arguments, `x` and `y`, which should each be a vector of
numbers. By default, the function will conduct a two-sample t-test assuming
unequal variances, which is what we want. We already have a vector for `x` in
`setosa_petal_length`. To get the `y` vector we do the same thing as above:

```{r}
# filter versicolor data
versicolor_data <- filter(iris, Species == "versicolor")

# extract petal lengths for versicolor
versicolor_petal_length <- pull(versicolor_data, Petal.Length)
```

Now we can use the two vectors in the t-test:

```{r}
# compare setosa and versicolor petal length
t.test(x = setosa_petal_length, y = versicolor_petal_length)
```

In the t-test results you can see the t-statistic, degrees of freedom,
p-value, and the alternative hypothesis.

Given that the p-value here is much less than 0.05 (it is very close to zero 
actually), we can reject the null hypothesis and accept the alternative
hypothesis by default, the difference in means is not equal to 0. Or phrased
phrased another way, the two means are not equal, they are different.

The t-test results go a step further and also give you the estimates of the
means for each species (1.462 and 4.260). To get the estimate of the difference
in petal length, you could subtract one number from the other. Here we
can see that *versicolor* petals are on average 2.798 mm longer than *setosa*
petals.

The results also give you a 95% confidence interval for that difference. Here we
can see that the confidence interval for the difference is about 2.66 to 2.94 
mm. Does that range include zero? No it does not, which is another way to
confirm that the difference in means is not equal to zero just as the hypothesis
test found.

## Assignment

For the assignmet we will use the `gapminder` data set in the **dslabs** 
package.


First, install **dslabs** using the Packages tab in RStudio. Then load the 
`gapminder` data set and turn it into a tibble so it prints nicely.

```{r, results=FALSE}
library(dslabs)
data("gapminder")
g <- as_tibble(gapminder)
g
```

Here is an interactive version of the table you can use to look at it. Note that
each country is grouped into regions and continents. You can sort the data by
clicking on the arrows next to column names, and filter by clicking in the box
below a column name.

```{r echo=FALSE}
DT::datatable(g, filter = "top", options = list(pageLength = 5))
```


For this assignment, which is a combination of lab 6 and a lab practical, I want 
you to demonstrate your data wrangling abilities to compare the means of groups.

Your assignment is to:

1. Calculate the mean infant mortality for each continent in the year 2000.
2. Plot the distribution of life expectancies on each continent in the year 2000.
3. Conduct a t-test to compare the mean fertility in the year 2000 in the 
continent "Americas" and another continent of your choice. For example, you 
could compare the Americas and Europe, or the Americas and Africa. Show your 
code and state the following:
    a. What are your null and alternate hypotheses?
    b. Do the means fertility rates differ?
    c. What is the 95% confidence interval for the difference in mean fertility
    rate?

## Template

Copy and paste this template into your R script.

Fill in the missing elements under the "assignment" heading

```{r, eval=FALSE}
#### Lab 6 / Lab Practical Script Template

### Practice ----------------------------------------------

## Objective 1. Visualize a normal distribution -----------

# turn iris into a tibble
library(tidyverse)
iris <- as_tibble(iris)
iris

# look at the dataset
help("iris")
glimpse(iris)
count(iris, Species)

# density plot of petal lengths by species
ggplot(iris, aes(x = Petal.Length, color = Species)) +
  geom_density()

# filter setosa data
setosa_data <- filter(iris, Species == "setosa")

# histogram of petal lengths for setosa
ggplot(setosa_data, aes(x = Petal.Length)) +
  geom_histogram(bins = 9)

# extract petal lengths for setosa
setosa_petal_length <- pull(setosa_data, Petal.Length)
setosa_petal_length

# calculate mean petal length for setosa
mean(setosa_petal_length)

# filter versicolor data
versicolor_data <- filter(iris, Species == "versicolor")

# extract petal lengths for versicolor
versicolor_petal_length <- pull(versicolor_data, Petal.Length)

# compare setosa and versicolor petal length
t.test(x = setosa_petal_length, y = versicolor_petal_length)


### Assignment -----------------------------------------
### ENTER YOUR CODE AND ANSWERS BELOW

# 1. Calculate the mean infant mortality for each continent in the year 2000.



# 2. Plot the distribution of life expectancies on each continent in 
#    the year 2000.



# 3. Conduct a t-test to compare the mean fertility in the year 2000
#    Compare the Americas with another continent of your choice

# 3a. Alternative hypothesis  Ha = 
#     Null hypothesis         H0 = 

# 3b. Do the means fertility rates differ? 



# 3c. What is the 95% confidence interval for the difference in mean 
#     fertility rate?



```

